### 2015-07-31 paper/paper.Rmd
few small fixes in the paper

What Can We Learn About Education Around the World from the PISA Data?
========================================================

@@

The OECD PISA Results in Focus report describes the survey as ``the world's global metric for quality, equity and efficiency in school education". The goal of the PISA survey is to assess the workforce readiness of 15-year old students. Nearly 500,000 students were tested across 65 countries and economies. Students were examined on how well they can apply the knowledge they learned in school to applications outside of school. The  reported scores range between 0-1000. Information about the students, parents, and schools is also collected. The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affect their performance in school. School principals responded to a questionnaire covering their school system and learning experiences for their students. In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.

@@

For each country, average scores on math and reading are calculated separately for boys and girls. These averages are differenced to compute the "gap" between boys and girls. To gauge the statistical significance of the difference, t-statistics and bootstrap confidence intervals were calculated. The results are plotted in the graphs below. Each dot corresponds to a difference between girls' and boys' scores, for each country, with lines giving the confidence interval. Wider intervals indicate less data support the statistics, for example, Liechtenstein tests few students so there is more uncertainty about the mean. Color indicates statistical significance (blue in favor of boys, pink in favor of girls, and green means there was no difference). The maps color the countries by the presence (or absence) of a gender gap.

The countries are sorted from biggest gap to smallest gap. Most countries have a gender gap in math scores in favor of boys. Colombia has the biggest gender gap, about 30 points on average boys do better than girls. It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.

Five countries have gender gaps in favor of girls; these are middle-eastern countries. Given popular perception of gender issues in the region, this may be a surprising result. Perhaps this is an example of how data can be used to show that reality is more nuanced than popular perception.

@@

For each country, we have computed the average math score by age that the children started school. In addition, for each country we computed the median age that children in the study started school. There are differences in typical school start age between countries. Ireland has the youngest median start age - children head to school at age 4. Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5. For many other countries the median start age is 6, and for northern Europe, much of Asia, and Brazil median start age is as late as 7. Some countries appear to have rigid starting ages, e.g. in Japan all children start at age 6.

Within a country, looking at the average math score for each start age, the score typically declines, especially after age 6. This is almost uniformly the pattern across all of the countries in the study. For several countries - Russia, Denmark, Bulgaria - if a child starts school at age 4 their math scores were much better on average than any other age group. School start age matters. Children need to start school in the 4-6 age range to expect better math scores, on average.

@@

The vast majority of students spend less than 10 hours studying per week. The pattern  suggests decreasing benefits to increased out of school study time after 10 hours. The thickness of the standard-error bars in the plot are an indication of the sample size.

@@

# 4. Parents matter!
### a. Number of parents in the home
The role of parents in educational achievement is the subject of much research. Looking at the number of parents in a household shows that students in two-parent households are, on average, performing better in math than students not in a two-parent households. The majority of students live in two-parent households. Of the single-parent households, a single mother is more common than a single father. Scores of students with just a mother at home tend to be higher than those with just a father at home. In many countries, mostly Asian, students from single-mother households achieve comparably to those from two-parent households. Students in households with no parents tend to perform the worst. This pattern is consistent across all countries.

```{r par_athomebar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE}
athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
athome$parents <- factor(athome$parents, levels=c("Both", "Mother", "Father", "Neither"))
#athome$name <- reorder(athome$name,athome$PV1MATH,mean, decreasing=F)
pbarhome <- qplot(x = parents, data = athome, xlab="Parents in home") 
pbarhome
```

@@

### b. Parent's Occupation Status
One of the variables the survey addressed was the occupation status of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. stay-at-home, retired)”, and no response. The frequencies for these categories can be seen in the bar chart. The chart shows that fathers primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are categorized as "Other". With women making up a larger portion of the work-force, it is interesting to look at how the parents' occupations relate to students' performance in math.

@@

Albania appears to be an outlier in many educational trends.  Many socioeconomic and school factors that play a role in other nations' test scores do not appear to affect Albania's test scores.  For example, Albania has one of the smallest gender gaps for math scores.  Additionally, number of possessions and truancy rates affected test scores for every nation except Albania.  This most likely indicates arrors in the data collected for the country.

@@

Generally, students with more possessions score higher on average.  The number of possessions is a surrogate for a student's economic standing.  The survey records whether students have desks, computers, internet, books, a place to study and dozens of other possessions.  After aggregating these number of possessions together, to get a total, we can see that scores increase on average as possessions increase in all nations, except Albania.

@@

@@

We have conducted what would be considered to be an exploratory analysis of the PISA data. There is a big difference between exploratory analyses and confirmatory analyses. When many people think of statistics they have in mind confirmatory methods, which involve deciding on a hypothesis, collecting the data, and doing a statistical test of significance. Exploratory data analysis is different, it focuses on the data at hand, and the goal is to let the data inform, to use various methods, primarily visual to find relationships and patterns and anomalies that may not have been anticipated ahead of the data collection. Exploratory analyses often lead to designed experiments where conditions are controlled, or specific data collection with all the controls to allow for hypothesis testing. Exploratory analysis is particularly important for today's big data society. Data is available everywhere, and typically collected because it is possible, without thought to testing particular hypotheses.

The PISA data is a very large data set, with 500,000 students measured on more than 600 variables, with additional information on schools and home life. The selection of students into the study follows a carefully designed survey instrument, which means that each student also has a sample weight which reflects how their demographic is represented in the sample. These allow the inference of findings made from this sample of students to the entire student population in the country. Strictly speaking, for the PISA data the survey weights primarily apply to the test score components, covering different types of schools and ensuring good representation of both genders. There is no guarantee that all socioeconomic groups are represented equally, or that households with all sorts of possible possessions are represented. The sampling design focused on school types and gender, only.

@@

The PISA data is a carefully designed survey, but the student and household questionaires are more loosely administered, yielding many missing values, and perhaps some questionable data. When results are surprising, and contradict current thinking, it can be helpful to research the problem in more details. For example, in follow up exploration of the PISA data during an introductory statistics class, we found that South Koreans do not use the internet as much as students in other countries. This is surprising, given the high tech nature of the society, and leads us to be skeptical about the data. There are several plausible reasons: (1) the mobile phone network dominates internet access more than internet providers, (2) individual identification on the questionaires might lead students not to answer honestly. If it is the latter, the problem could be fixed by re-designing the survey instrument.

Our findings about test scores and other factors, documented in this article, some results that are consistent with current thinking, e.g. parents matter, socioeconomic status matters, time spent studying matters, but also some results that surprised us. These findings are what this data suggests, but they are only as good as the data. If the collection was not well done, and the collection is the responsibility of the local committee in each country, then the findings can not be trusted. An example of this is the data from Albania, which our analysis suggests that the PISA committee needs to re-assess what was done by that local committee, because the results from the questionaire are at odds with all other countries.

Overall, our analysis yielded several interesting insights into factors that affect the average test scores. The data is rich, and should reliably reflect the larger population of 15 year olds. It is reassuring, though, that repeatedly we found that the major source of variation in the data is from individual to individual. For example, even though, on average boys perform better on the math tests in the USA, and girls on the reading tests, the top math score was achieved by a girl, and the top reading score was achieved by a boy. You as a person are more than your demographic -  you count, your individual efforts matter.

@@

The following R packages were used to conduct the data exploration: shiny, plyr, dplyr, ggplot2, stringr, reshape2, lubridate, maps, ggmap, htmltools, rworldmap, grid, scales, doBy. Principles of reproducible research were used for the analysis and the markdown script is available from https://github.com/ijlyttle/isu_pisa/blob/master/paper/paper.Rmd.

>>>>

---
output: word_document
---
What Does the Data Collected During PISA Testing of Teenagers Tell Us About Education Around the World?
========================================================

@@

The OECD PISA Results in Focus report describes the survey as ``the world's global metric for quality, equity and efficiency in school education". The goal of the PISA survey is to assess the workforce readiness of 15-year old students. Nearly 500,000 students were tested across 65 countries and 18,000 schools. Students were examined on how well they can apply the knowledge they learned in school to applications outside of school. The  reported scores range between 0-1000. Information about the students, parents, and schools is also collected. The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affect their performance in school. School principals responded to a questionnaire covering their school system and learning experiences for their students. In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.

@@

For each country, average scores on math and reading are calculated separately for boys and girls. These averages are differenced to compute the "gap" between boys and girls. To gauge the statistical significance of the difference, t-statistics and bootstrapped 95% confidence intervals were calculated. The results are plotted in the graphs below. Each dot corresponds to a difference between girls' and boys' scores, for each country, with lines giving the confidence interval. Wider intervals indicate less data support the statistics, for example, Liechtenstein tests few students so there is more uncertainty about the mean. Color indicates statistical significance (blue in favor of boys, pink in favor of girls, and green means there was no difference). The maps color the countries by the presence (or absence) of a gender gap.

The countries are sorted from biggest gap to smallest gap. Most countries have a gender gap in math scores in favor of boys. Colombia has the biggest gender gap, about 30 points, on average boys do better than girls. It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.

Five countries have gender gaps in favor of girls; these are middle-eastern countries, Malaysia and Thailand. Given popular perception of gender issues in the Middle East, this may be a surprising result. Perhaps this is an example of how data can be used to show that reality is more nuanced than popular perception.

@@

For each country, we have computed the average math score by age that the students started school. In addition, for each country we computed the median age that children in the study started school. There are differences in typical school start age between countries. Ireland has the youngest median start age - children head to school at age 4. Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5. For many other countries the median start age is 6, and for northern Europe, much of Asia, and Brazil median start age is as late as 7. Some countries appear to have rigid starting ages, e.g. in Japan all children start at age 6.

Within a country, looking at the average math score for each start age, the score typically declines, especially after age 6. This is almost uniformly the pattern across all of the countries in the study. For several countries - Russia, Denmark, Bulgaria - if a child starts school at age 4 their math scores were much better on average than any other age group. School start age appears to matter: with starting school in the 4-6 age range being associated with higher average math scores. There could be many other factors confounding this observation, most notably being that it is the more mathematically gifted students that start earlier, leading to a difference in the population characteristics of the different age (at school start) groups.

@@

The vast majority of students spend less than 10 hours studying per week. The pattern  suggests decreasing benefits to increased out of school study time after 10 hours. The thickness of the standard-error bars in the plot are an indication of the sample size, with thicker line indicating less certainty about the math value.

@@

# 4. Parents matter!
### a. Number of parents in the home
The role of parents in educational achievement is the subject of much research. Looking at the number of parents in a household shows that students in two-parent households are, on average, performing better in math than students not in a two-parent households. The majority of students (97%) live in two-parent households. Of the single-parent households, a single mother is five times more common than a single father. Scores of students with just a mother at home tend to be higher than those with just a father at home. In many countries, mostly Asian, students from single-mother households achieve comparably to those from two-parent households. Students in households with no parents tend to perform the worst. This pattern is consistent across all countries.

```{r par_athomebar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE}
athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
athome$parents <- factor(athome$parents, levels=c("Both", "Mother", "Father", "Neither"))
#athome$name <- reorder(athome$name,athome$PV1MATH,mean, decreasing=F)
#pbarhome <- qplot(x = parents, data = athome, xlab="Parents in home") 
#pbarhome
```

@@

### b. Parent's Occupation Status
One of the variables the survey addressed was the occupation status of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. stay-at-home, retired)”, and no response. Fathers primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are categorized as "Other". With women making up a larger portion of the work-force, it is interesting to look at how the parents' occupations relate to students' performance in math.

@@

Albania appears to be an outlier in many educational trends.  Many socioeconomic and school factors that are associated with test scores in other nations are unrelated to Albania's test scores.  For example, there is no difference in math scores based on household structure in Albania, or in terms of hours spent studying.

@@

The lack of association where we would expect association is also evident in number of possessions. Generally, students with more possessions score higher on average.  The number of possessions is a surrogate for a student's economic standing.  The survey records whether students have desks, computers, internet, books, a place to study and dozens of other possessions.  After aggregating these number of possessions together, to get a total, we can see that scores increase on average as possessions increase in all nations, except Albania.

@@

These patterns, especially all together, would most likely indicate that there are some problems with the way the data was collected for Albania.

@@

The PISA data is a very large data set, with 500,000 students measured on more than 600 variables, and additional information collected about school and home life. The selection of students into the study follows a carefully designed survey instrument, which means that each student also has a sample weight which reflects how their demographic is represented in the sample. These allow the inference of findings made from this sample of students to the entire student population in the country. Strictly speaking, for the PISA data the survey weights primarily apply to the test score components, covering different types of schools and ensuring good representation of both genders. There is no guarantee that all socioeconomic groups are represented equally, or that households with all sorts of possible possessions are represented. The sampling design focused on school types and gender, only.

We have conducted what would be considered to be an exploratory analysis of the PISA data. There is a big difference between exploratory analyses and confirmatory analyses. When many people think of statistics they have in mind confirmatory methods, which involve deciding on a hypothesis, collecting the data, and doing a statistical test of significance. Exploratory data analysis is different, it focuses on the data at hand, and the goal is to let the data inform, to use various methods, primarily visual to find relationships and patterns and anomalies that may not have been anticipated ahead of the data collection. Exploratory analyses often lead to designed experiments where conditions are controlled, or specific data collection with all the controls to allow for hypothesis testing. Exploratory analysis is particularly important for today's big data society. Data is available everywhere, and in many situations it is collected because it is easy to. Mining data, like we have done for the PISA survey data, can lead to constructing hypotheses that can be rigorously tested with new data.

@@

The PISA data is a carefully designed survey, but the student and household questionaires are more loosely administered, yielding many missing values, and perhaps some questionable data. When results are surprising, and contradict current thinking, it can be helpful to research the problem in more details. For example, in follow up exploration of the PISA data during an introductory statistics class, we found that South Koreans do not use the internet as much as students in other countries. This is surprising, given the high tech nature of the society, and leads us to be skeptical about that part of the data. There are several plausible reasons: (1) the mobile phone network dominates internet access more than internet providers, and so internet usage is masked by phone usage, (2) individual identification on the questionaires might lead students not to answer honestly. Both  problems could be fixed by re-designing the survey instrument.

Our findings about test scores and other factors, documented in this article, some of the results are consistent with current thinking, e.g. parents matter, socioeconomic status matters, time spent studying matters, but also contained some results that surprised us. These findings are what this data suggests, but they are only as good as the data. If the collection was not well done, and the collection is the responsibility of the local committee in each country, then the findings cannot be trusted. An example of this is the data from Albania, which our analysis suggests that the PISA committee needs to re-assess what was done by that local committee, because the results from the questionaire are at odds with all other countries.

Overall, our analysis yielded several interesting insights into factors that affect the average test scores. The data is rich, and should reliably reflect the larger population of 15 year olds. Average scores are not the individual experience.  We found that the major source of variation in the data is from individual to individual. For example, even though, on average boys perform better on the math tests in the USA, and girls on the reading tests, the top math score was achieved by a girl, and the top reading score was achieved by a boy. You as a person are more than your demographic -  you count, your individual efforts matter.

@@

The following R packages were used to conduct the data exploration: shiny, plyr, dplyr, ggplot2, stringr, reshape2, lubridate, maps, ggmap, htmltools, rworldmap, grid, scales, doBy, boot. Principles of reproducible research were used for the analysis and the markdown script is available from https://github.com/ijlyttle/isu_pisa/blob/master/paper/paper.Rmd. By downloading the data from the OECD site, all of our results can be reproduced in the free, and open source, software R, particularly using the RStudio interface.
